---
title: "Cause_of_Death"
author: "Jorge Alejandro, Mia Alanis, Bryce Borak"
output: 
  html_document:
    toc: true
    code_folding: show
  
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(shiny)
library(knitr)
library(ggplot2)
library(kableExtra)
library(stringr)
library(sf) # For map handling
library(rnaturalearth) # For getting the Europe map outlines
install.packages("viridis", repos = "https://cran.rstudio.com/")
library(viridis) # For the color palette
install.packages("ggrepel", repos = "https://cran.rstudio.com/")
library(ggrepel) #best for labels
```


```{r, message=FALSE, warning=FALSE, include=FALSE}
#setting up the datasets we will be using for our projects
  death_ds <- read.csv("cause_of_deaths.csv")
  popu_ds <- read.csv("world_population.csv")

#Im going to add other columns that include things such as "Total recorded deaths" and a calculation for the crude death rate.
  death_ds %>% mutate(Total.Death = rowSums(death_ds[, -c(1:3)], na.rm = TRUE)) -> death_ds
  
  death_ds %>% 
    filter(Year %in% c(1970, 1980, 1990, 2000, 2010, 2015, 2020, 2022)) -> death_ds
  
#This changes the population dataset to a viable format that we could use to join the death and the population datasheets
  popu_ds %>% 
    pivot_longer(
      cols = matches("^X\\d"),
      names_to = "Year",
      values_to = "Population"
    ) %>% 
    mutate(Year = as.numeric(gsub("X|\\.Population", "", Year))) -> popu_longer
  
#For a comprehensive dataset with population added to it, I made a join so we can analyse it further.
  dth_pplt_ds <- inner_join(death_ds, popu_longer, join_by(Country.Territory, Year))
  
#We might also want to use the numbers of the death rate of a population so im going to mutate the dataset to have that in the final column
  
  dth_pplt_ds %>% mutate(Crude.Death.Rate = (Total.Death / Population) * 100, .after = "Total.Death") -> dth_pplt_ds

```

## Our Data
  
  Our data is sourced from Kaggle from a user named "Sourav Banerjee" who is a data scientist who works at R systems. On Kaggle, he has the title of "Grandmaster". This was the main reason we picked his two data sets because of his credibility and experience to format and clean data sets. 
  
  The two data sets that we picked for our project is "Cause of Deaths" and "World Population." Our main focus is the "Cause of Deaths" data set. which goes into different causes of deaths around the world. These deaths range from deaths due to natural causes such as diseases and chronic health issues to death due to accidents such as drowning or fires. This data set gives a number of total deaths due to a given cause within a given year. A long with this information, it also gives us information such as what countries are affected by these causes.
  
  Our main purpose for the "World Population" data set is for us to be able to cross refrense this data so that we can get more information of the country as well as get information for a continent.
  
::: {style="float: left; width: 45%; border-right: 2px solid steelblue; padding-right: 20%;"}
### **Raw Input**

**World Population** \* 234 Rows \| 17 Columns \* [Source Link](https://www.kaggle.com/datasets/iamsouravbanerjee/world-population-dataset)

**Causes of Death** \* 6,120 Rows \| 34 Columns \* [Source Link](https://www.kaggle.com/datasets/iamsouravbanerjee/cause-of-deaths-around-the-world)
:::

::: {style="float: right; width: 45%; color: midnightblue;"}
### **Final Data Set**

-   **796** Rows
-   **45** Columns

**Key Operations:** 1. **Filter:** Decadal years (1970-2022) 2. **Pivot:** Population into long format 3. **Join:** Inner join by Country/Year 4. **Mutate:** Added Crude Death Rate
:::

```{r, echo=FALSE, fig.width=8, fig.height=15}
dth_pplt_ds %>%
  head(100) %>%
  # Remove the comma from Year and format numbers
  mutate(Year = as.character(Year),
         across(where(is.numeric), ~scales::comma(.))) %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("bordered", "condensed", "hover"), 
    full_width = FALSE,  # This keeps the table from stretching
    position = "center",  # Centers the smaller table on the page
    font_size = 18    # Manually shrink the text size (default is usually 16)
  ) %>%
  row_spec(0, background = "steelblue", color = "white") %>%
  # Explicitly set column widths to keep it "thin"
  column_spec(1:ncol(dth_pplt_ds), border_left = TRUE, border_right = TRUE) %>% 
  scroll_box(width = "1400px", height = "350px") # Set a fixed pixel width instead of 100%
```

## What is Shown in our Data Set

### What Makes up the Majority of Our Columns?

Our dataset show various different death causes, both natural and unnatural.

```{r, echo = FALSE, warning = FALSE, fig.width=8, fig.height=13 }
all_cause_cols <- colnames(dth_pplt_ds)[4:34]

non_diseases <- c("Drowning", "Interpersonal.Violence", "Drug.Use.Disorders", "Alcohol.Use.Disorders", "Self.harm", "Exposure.to.Forces.of.Nature", "Environmental.Heat.and.Cold.Exposure", "Conflict.and.Terrorism", "Road.Injuries", "Poisonings", "Fire..Heat..and.Hot.Substances", "Maternal.Disorders")

clean_non_diseases <- str_replace_all(non_diseases, "\\.", " ")

disease_cols <- setdiff(all_cause_cols, non_diseases)

clean_diseases_cols <- str_replace_all(disease_cols, "\\.", " ")

#this is just separating the columns with out the death types
other_columns <- setdiff(colnames(dth_pplt_ds), all_cause_cols)

clean_other_cols <- str_replace_all(other_columns, "\\.", " ")

other_cols_table <- data.frame(
  Other_Columns = clean_other_cols
)

max_ln_all <- max(length(clean_other_cols), 
                  length(clean_diseases_cols), 
                  length(clean_non_diseases))

#this is creating the combined data frame
combined_dictionary <- data.frame(
  Other_Columns = c(clean_other_cols, rep("", max_ln_all - length(clean_other_cols))),
  Disease_Causes = c(clean_diseases_cols, rep("", max_ln_all - length(clean_diseases_cols))),
  Non_Disease_Causes = c(clean_non_diseases, rep("", max_ln_all - length(clean_non_diseases)))
)

#this is to create a table to show both diseases and non-diseases but also the other columns
max_ln <- max(length(clean_diseases_cols) , length(clean_non_diseases))

comparison_table <- data.frame(
  Disease_Causes = c(clean_diseases_cols, rep(NA, max_ln - length(clean_diseases_cols))),
  Non_Disease_Causes = c(clean_non_diseases, rep(NA, max_ln - length(clean_non_diseases)))
)

comparison_table[is.na(comparison_table)] <- ""

#this create a better looking Table
combined_dictionary %>%
  kbl(col.names = c("Data Columns", "Diseases Columns", "Non-Disease Columns"),
      caption = "Dataset Column Classification",
      align = "lll") %>%
  kable_styling(bootstrap_options = c("bordered", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 26) %>%
  # Header Styling
  row_spec(0, background = "steelblue", color = "white", bold = TRUE) %>%
  # Column Boxing and Widths
  column_spec(1, border_left = TRUE, border_right = TRUE, width = "15em", bold = TRUE) %>%
  column_spec(2, border_right = TRUE, width = "18em") %>%
  column_spec(3, border_right = TRUE, width = "18em") %>% 
  scroll_box(width = "1400px", height = "550px")
```

------------------------------------------------------------------------

::: {style="float: left; width: 48%; padding-top: 10px;"}
### **The Data Collaboration**

The **"Cause of Deaths"** dataset provides the *counts*, but the **"World Population"** dataset provides the *context*.

By merging them, we move from raw totals to meaningful **rates**. This allows us to answer not just *"how many died?"*, but **"how heavily was the population actually impacted?"**

<br>

### **The Investment Perspective**

**Imagine we are a global healthcare fund.** How do we choose where to invest? 

We don't just look at who has the most deaths; we analyze:
- **The Crude Death Rate:** The true burden on the people.
- **The Trend:** Is the crisis improving or worsening?
:::

::: {style="float: right; width: 45%; background-color: #F0F8FF; padding: 10px; border-radius: 12px; border: 2px solid #4682B4; color: #191970;"}
### **Our Core Questions**

- **Regional Burden:** Which continents face the heaviest impact from biological diseases?


- **Distribution & Trends:**
How do specific diseases distribute across the continent?


- **The Cardiovascular Divide:**
Which nations face the highest mortality burden from heart disease?

:::


```{r echo=FALSE, warning = FALSE}
## Continents with the Most Deaths

ntrl_deaths <- dth_pplt_ds %>% 
  select(-c(all_of(non_diseases), Total.Death, Crude.Death.Rate))

ntrl_deaths <- ntrl_deaths %>% 
  mutate(
    Total.Deaths = rowSums(across(all_of(disease_cols)), na.rm = TRUE), 
    # Standard crude rate is usually per 1,000 or 100,000
    # Your current (Total/Pop)/100 is essentially a percentage.
    Crude.Death.Rate = (Total.Deaths / Population) * 100000
  )
```

---

## Regional Burden

### Which continents face the heaviest impact from biological diseases?

```{r, echo=FALSE}
ntrl_deaths <- ntrl_deaths %>% 
  select(-c(Code, Rank, CCA3, Capital, Density..per.km.., Growth.Rate, World.Population.Percentage, Area..km..))  %>% 
  relocate(Continent, .after = Country.Territory) %>% 
  relocate(Total.Deaths, .after = Acute.Hepatitis)

#Summarize with the Weighted Rate
# We save this to a new object for your plot/table
dis_crude_rate <- ntrl_deaths %>% 
  group_by(Continent, Year) %>%
  summarize(
    # We must SUM the deaths and population for the continent
    Total_Deaths_Continent = sum(Total.Deaths, na.rm = TRUE),
    # Calculate the rate here so the next line can 'arrange' by it
    Crude.Death.Rate = (Total_Deaths_Continent / sum(Population, na.rm = TRUE)) * 100000,
    .groups = "drop"
  ) %>%
  # Now arrange works because Crude.Death.Rate exists in this table
  arrange(Year, desc(Crude.Death.Rate))

#disease crude death rate
dis_crude_rate <- ntrl_deaths %>%
  group_by(Continent, Year) %>%
  summarize(
    # name it clearly to distinguish it from the overall Crude Death Rate
    Avg_Dis_Death_Rate = mean((Total.Deaths / Population) * 100000, na.rm = TRUE)
  , .groups = "drop" ) %>%
  arrange(desc(Avg_Dis_Death_Rate))

```

::: {.columns}

::: {.column width="65%"}
```{r, echo=FALSE}
# --- STEP 2: GRAPH RENDER ---
ggplot(dis_crude_rate, aes(x = Year, y = Avg_Dis_Death_Rate, color = Continent)) +
  geom_line(linewidth = 2.5) +
  geom_point(size = 4) +
  # Highlighting Europe with a specific color if desired
  scale_color_brewer(palette = "RdYlBu") +
  theme_minimal() +
  labs(title = "The Mortality Trajectory",
       subtitle = "Disease-driven death rates per 100k (1990 - 2015)",
       x = "Year",
       y = "Deaths per 100,000") +
  theme(
    plot.title = element_text(face = "bold", color = "#191970", size = 20),
    legend.position = "bottom",
    # MAKE AXIS WORDING BIGGER
    # Axis Titles ("Year" and "Deaths per 100,000")
    axis.title = element_text(size = 16, face = "bold"), 
    
    # 2. Axis Text (The actual numbers/years on the scales)
    axis.text = element_text(size = 14, color = "black"),
    
    # 3. Legend Text (The continent names)
    legend.text = element_text(size = 12)
  )
```
:::

::: {.column width="35%"}
<style>
.table {
    margin-top: 30px; /* Moves table down from the header */
    border: 2px solid #4682B4; /* Adds a Steelblue border around the whole table */
}
</style>
```{r, echo=FALSE}
# --- STEP 3: TABLE RENDER ---
ntrl_deaths %>%
  filter(Year %in% c(1990, 2015)) %>%
  group_by(Continent, Year) %>%
  summarize(
    Avg_Rate = (sum(Total.Deaths, na.rm = TRUE) / sum(Population, na.rm = TRUE)) * 100000,
    .groups = "drop"
  ) %>%
  # This "spreads" the years into their own columns
  pivot_wider(names_from = Year, values_from = Avg_Rate) %>%
  mutate(Growth = `2015` - `1990`) %>% # Optional: Shows the change
  arrange(desc(`2015`)) %>%
  kable(digits = 1, col.names = c("Continent", "1990 Rate", "2015 Rate", "Absolute Change")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 18) %>%
  # Highlight the 2015 column to show it's the "Current" status
  column_spec(3, bold = TRUE, background = "#F0F8FF") %>%
  row_spec(0, background = "steelblue", color = "white")

```
<div style="background-color: #ffffff; padding: 15px; border-left: 5px solid #191970;color: #191970; font-size: 0.9em; margin-top: 15px; box-shadow: 2px 2px 5px rgba(0,0,0,0.1);">
We show deaths per 100,000 people so that large countries don't overshadow smaller ones. This reveals when major outbreaks or successful medical breakthroughs shifted the numbers.
</div>
:::

:::

---

## Distribution & Trends: Europe Deep-Dive

### How do specific diseases distribute across the continent?

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# DATA PREPARATION
# Clean names and filter for Europe once
europe_base <- ntrl_deaths %>%
  filter(Continent == "Europe") %>%
  select(all_of(disease_cols), Country.Territory, Year, Population) %>%
  pivot_longer(cols = all_of(disease_cols), names_to = "Disease", values_to = "Deaths") %>%
  mutate(Disease = str_replace_all(Disease, "\\.", " "))

# Prepare Table Data (Count countries)
europe_summary_table <- europe_base %>%
  filter(Deaths > 0) %>%
  group_by(Disease) %>%
  summarize(Countries = n_distinct(Country.Territory), .groups = "drop") %>%
  arrange(desc(Countries))

# Prepare Graph Data (Top diseases > 40 countries)
top_diseases <- europe_summary_table$Disease[europe_summary_table$Countries > 40]

graph_data <- europe_base %>%
  filter(Disease %in% top_diseases) %>%
  group_by(Year, Disease) %>%
  summarise(Total_Deaths = sum(Deaths, na.rm = TRUE), .groups = "drop")
```

::: {.columns}

::: {.column width="35%"}

<style>
/* This fixes the table border you wanted */
.scrollbox-area {
border: 2px solid #4682B4;
margin-top: 10px;
}
</style>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#TABLE RENDER
europe_summary_table %>%
  kable(format = "html", col.names = c("Cause of Death", "Countries Affected")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 16) %>%
  row_spec(0, background = "steelblue", color = "white") %>%
  scroll_box(height = "400px", width = "100%")
```

<div style="background-color: #ffffff; padding: 12px; border-left: 5px solid #191970; color: #191970; font-size: 0.8em; margin-top: 10px; box-shadow: 2px 2px 5px rgba(0,0,0,0.1); line-height: 1.4;">
  <strong>Key Insight:</strong> This view highlights the "Widespread" paradox. While many diseases affect every European nation, the logarithmic trends show a steady decline in communicable deaths while chronic conditions remain stable.
</div>

:::

::: {.column width="65%"}

```{r, echo=FALSE, fig.width=9, fig.height=6, message=FALSE, warning=FALSE}
# We find the top 11 diseases based on the most recent year's totals
top_11_diseases <- graph_data %>%
  filter(Year == max(Year)) %>%
  arrange(desc(Total_Deaths)) %>%
  slice_head(n = 11) %>%
  pull(Disease)

# Filter the main dataset to only include these 11
graph_data_top11 <- graph_data %>%
  filter(Disease %in% top_11_diseases)

# GRAPH RENDER 
ggplot(graph_data_top11, aes(x = Year, y = Total_Deaths, color = Disease)) +
  geom_line(linewidth = 1.2, alpha = 0.8) +
  geom_point(size = 2) + 
  scale_y_log10() + # Added comma labels for readability
  
  # The RdYlBu palette has exactly 11 colors
  scale_color_brewer(palette = "RdYlBu") + 
  
  theme_minimal() +
  labs(title = "Mortality Trends in Europe",
       subtitle = "Top 11 Causes of Death (Logarithmic Scale)",
       x = "Year",
       y = "Total Deaths",
       color = "Cause of Death") +
  theme(
    plot.margin = margin(t = 10, r = 20, b = 10, l = 20), # Increased right margin
    plot.title = element_text(face = "bold", color = "#191970", size = 20),
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12, color = "black")
  ) + 
  # Adjusted to fit 11 items neatly (2 columns)
  guides(color = guide_legend(ncol = 2, byrow = TRUE))
```

:::

:::
---

## The Cardiovascular Divide: The Cardiovascular Landscape of Europe (2015)

### Which nations face the highest mortality burden from heart disease?

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# map data for Europe
europe_map <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf")

#Filter your data for Cardio Deaths in the most recent year (2015)
cardio_2015 <- ntrl_deaths %>%
  filter(Continent == "Europe", Year == 2015) %>%
  mutate(Country.Territory = case_when(
    Country.Territory == "Russian Federation" ~ "Russia",
    Country.Territory == "United Kingdom" ~ "United Kingdom",
    Country.Territory == "Czech Republic" ~ "Czechia",
    TRUE ~ Country.Territory  # This keeps all other names the same
  )) %>%
  # We use Country.Territory since Code is missing
  select(Country.Territory, Population, any_of("Cardiovascular.Diseases")) %>%
  mutate(Cardio_Rate_100k = (Cardiovascular.Diseases / Population) * 100000)

#join your data to the map data using the ISO Code (e.g., "GBR", "FRA")
europe_map_data <- europe_map %>%
  left_join(cardio_2015, by = c("name" = "Country.Territory")) %>%
  filter(continent == "Europe")  # Keeps the map focused

```

::: {.columns style="display: flex; align-items: flex-start;"}

::: {.column width="75%"}

```{r, echo=FALSE,fig.width=7, fig.height=5.5, out.width="100%", message=FALSE, warning=FALSE}

top_5_labels <- europe_map_data %>%
  filter(!is.na(Cardio_Rate_100k)) %>%
  arrange(desc(Cardio_Rate_100k)) %>%
  slice_head(n = 5)

# MAP RENDER (The "Big Visual")
ggplot(data = europe_map_data) +
  # Draw the country polygons and fill by the Cardio Rate
  geom_sf(aes(fill = Cardio_Rate_100k), color = "white", linewidth = 0.3) +
  geom_label_repel(
    data = top_5_labels, 
    aes(label = name, geometry = geometry),
    stat = "sf_coordinates",
    size = 3.5,
    fontface = "bold",
    color = "#191970",           # Dark blue text
    fill = alpha("white", 0.7),  # Semi-transparent white box for readability
    box.padding = 0.5,           # Keeps labels away from each other
    segment.color = 'black'     # Adds a small line pointing to the country
  ) +
  # Set the coordinate limits to zoom in on Europe (excluding most of Russia)
  coord_sf(xlim = c(-25, 70), ylim = c(34, 72), expand = FALSE) +
  
  # Define the Steel Blue color scale (from light to dark)
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "#F0F0F0",
                       name = "Deaths per 100k", labels = scales::comma) +
  
  theme_minimal() +
  labs(title = "Cardiovascular Mortality Rates",
       subtitle = "Deaths per 100,000 population",
       caption = "Note: 2015 Data. Russia included up to the Ural Mountains.") +
  
  theme(
    plot.title = element_text(face = "bold", color = "#191970", size = 22),
    plot.subtitle = element_text(size = 14, color = "black"),
    legend.position = "right", # Puts the key on the side
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    # Hide the background grid lines for a cleaner look
    panel.grid = element_blank(),
    axis.text = element_blank(), # Hide lat/long labels
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )
```
:::

::: {.column width="25%" style="margin-left: -2%;"}

<div style="margin-top: 40px;">

```{r, echo=FALSE,}
#  TOP 5 TABLE RENDER 
cardio_2015 %>%
  select(Country.Territory, Cardio_Rate_100k) %>%
  arrange(desc(Cardio_Rate_100k)) %>%
  kable(format = "html", digits = 1, col.names = c("Country", "Rate")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), 
                font_size = 13, 
                full_width = TRUE) %>% 
  row_spec(0, background = "#4682B4", color = "white", extra_css = "position: sticky; top: 0;") %>%
  # Reset width to 100% so it stays inside its column
  scroll_box(height = "280px", width = "100%")

```

</div>

<div style="background-color: #ffffff; padding: 10px; border-left: 4px solid #191970; color: #191970; font-size: 0.75em; margin-top: 10px; line-height: 1.3;">
<strong>Key Insight:</strong> The data highlights a stark East-West divide. High mortality rates cluster in Eastern Europe, while Western and Northern nations show a significantly lower burden.
</div>
:::

:::
