---
title: "Cause_of_Death"
author: "Jorge Alejandro, Mia Alanis, Bryce Borak"
output: slidy_presentation
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(shiny)
library(knitr)
library(ggplot2)
library(kableExtra)
library(stringr)

#setting up the datasets we will be using for our projects
  death_ds <- read.csv("cause_of_deaths.csv")
  popu_ds <- read.csv("world_population.csv")

#Im going to add other columns that include things such as "Total recorded deaths" and a calculation for the crude death rate.
  death_ds %>% mutate(Total.Death = rowSums(death_ds[, -c(1:3)], na.rm = TRUE)) -> death_ds
  
  death_ds %>% 
    filter(Year %in% c(1970, 1980, 1990, 2000, 2010, 2015, 2020, 2022)) -> death_ds
  
#This changes the population dataset to a viable format that we could use to join the death and the population datasheets
  popu_ds %>% 
    pivot_longer(
      cols = matches("^X\\d"),
      names_to = "Year",
      values_to = "Population"
    ) %>% 
    mutate(Year = as.numeric(gsub("X|\\.Population", "", Year))) -> popu_longer
  
#For a comprehensive dataset with population added to it, I made a join so we can analyse it further.
  dth_pplt_ds <- inner_join(death_ds, popu_longer, join_by(Country.Territory, Year))
  
#We might also want to use the numbers of the death rate of a population so im going to mutate the dataset to have that in the final column
  
  dth_pplt_ds %>% mutate(Crude.Death.Rate = (Total.Death / Population) * 1000, .after = "Total.Death") -> dth_pplt_ds

```

## Our Data

::: {style="float: left; width: 45%; border-right: 2px solid steelblue; padding-right: 20%;"}
### **Raw Input**

**World Population** \* 234 Rows \| 17 Columns \* [Source Link](https://www.kaggle.com/datasets/iamsouravbanerjee/world-population-dataset)

**Causes of Death** \* 6,120 Rows \| 34 Columns \* [Source Link](https://www.kaggle.com/datasets/iamsouravbanerjee/cause-of-deaths-around-the-world)
:::

::: {style="float: right; width: 45%; color: midnightblue;"}
### **Final Data Set**

-   **796** Rows
-   **45** Columns

**Key Operations:** 1. **Filter:** Decadal years (1970-2022) 2. **Pivot:** Population into long format 3. **Join:** Inner join by Country/Year 4. **Mutate:** Added Crude Death Rate
:::

```{r, echo=FALSE, fig.width=8, fig.height=15}
dth_pplt_ds %>%
  head(100) %>%
  # Remove the comma from Year and format numbers
  mutate(Year = as.character(Year),
         across(where(is.numeric), ~scales::comma(.))) %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("bordered", "condensed", "hover"), 
    full_width = FALSE,  # This keeps the table from stretching
    position = "center",  # Centers the smaller table on the page
    font_size = 18    # Manually shrink the text size (default is usually 16)
  ) %>%
  row_spec(0, background = "steelblue", color = "white") %>%
  # Explicitly set column widths to keep it "thin"
  column_spec(1:ncol(dth_pplt_ds), border_left = TRUE, border_right = TRUE) %>% 
  scroll_box(width = "1400px", height = "350px") # Set a fixed pixel width instead of 100%
```

## What is Shown in our Data Set

### What Makes up the Majority of Our Columns?

Our dataset show various different death causes, both natural and unnatural.

```{r, echo = FALSE, warning = FALSE, fig.width=8, fig.height=13 }
all_cause_cols <- colnames(dth_pplt_ds)[5:34]

non_diseases <- c("Drowning", "Interpersonal.Violence", "Drug.Use.Disorders", "Alcohol.Use.Disorders", "Self.harm", "Exposure.to.Forces.of.Nature", "Environmental.Heat.and.Cold.Exposure", "Conflict.and.Terrorism", "Road.Injuries", "Poisonings", "Fire..Heat..and.Hot.Substances", "Maternal.Disorders")

clean_non_diseases <- str_replace_all(non_diseases, "\\.", " ")

disease_cols <- setdiff(all_cause_cols, non_diseases)

clean_diseases_cols <- str_replace_all(disease_cols, "\\.", " ")

#this is just separating the columns with out the death types
other_columns <- setdiff(colnames(dth_pplt_ds), all_cause_cols)

clean_other_cols <- str_replace_all(other_columns, "\\.", " ")

other_cols_table <- data.frame(
  Other_Columns = clean_other_cols
)

max_ln_all <- max(length(clean_other_cols), 
                  length(clean_diseases_cols), 
                  length(clean_non_diseases))

#this is creating the combined data frame
combined_dictionary <- data.frame(
  Other_Columns = c(clean_other_cols, rep("", max_ln_all - length(clean_other_cols))),
  Disease_Causes = c(clean_diseases_cols, rep("", max_ln_all - length(clean_diseases_cols))),
  Non_Disease_Causes = c(clean_non_diseases, rep("", max_ln_all - length(clean_non_diseases)))
)

#this is to create a table to show both diseases and non-diseases but also the other columns
max_ln <- max(length(clean_diseases_cols) , length(clean_non_diseases))

comparison_table <- data.frame(
  Disease_Causes = c(clean_diseases_cols, rep(NA, max_ln - length(clean_diseases_cols))),
  Non_Disease_Causes = c(clean_non_diseases, rep(NA, max_ln - length(clean_non_diseases)))
)

comparison_table[is.na(comparison_table)] <- ""

#this create a better looking Table
combined_dictionary %>%
  kbl(col.names = c("Data Columns", "Diseases Columns", "Non-Disease Columns"),
      caption = "Dataset Column Classification",
      align = "lll") %>%
  kable_styling(bootstrap_options = c("bordered", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 26) %>%
  # Header Styling
  row_spec(0, background = "steelblue", color = "white", bold = TRUE) %>%
  # Column Boxing and Widths
  column_spec(1, border_left = TRUE, border_right = TRUE, width = "15em", bold = TRUE) %>%
  column_spec(2, border_right = TRUE, width = "18em") %>%
  column_spec(3, border_right = TRUE, width = "18em") %>% 
  scroll_box(width = "1400px", height = "550px")
```

------------------------------------------------------------------------

::: {style="float: left; width: 48%; padding-top: 20px;"}
### **The Data Collaboration**

The **"Cause of Deaths"** dataset provides the *counts*, but the **"World Population"** dataset provides the *context*.

By merging them, we move from raw totals to meaningful **rates**. This allows us to answer not just *"how many died?"*, but *"how heavily was the population actually impacted?"*
:::

::: {style="float: right; width: 45%; background-color: #F0F8FF; padding: 25px; border-radius: 12px; border: 3px solid #4682B4; color: #191970;"}
### **Our Core Questions**

-   **Regional Burden** Which continents face the heaviest impact from biological diseases?

-   **Fair Comparison** How do we level the playing field between small nations and global giants?

-   **Trends Over Time** How has the relationship between population growth and mortality evolved from 1990 to 2015?

-   **Global Reach** Which specific diseases are the most widespread across international borders?
:::

::: {style="clear: both;"}
:::

------------------------------------------------------------------------

```{r echo=FALSE, warning = FALSE}
## Continents with the Most Deaths

dth_pplt_ds %>% 
  select(-c(non_diseases, Total.Death, Crude.Death.Rate)) -> ntrl_deaths

ntrl_deaths %>% 
  mutate(Total.Deaths = rowSums(ntrl_deaths[, 4:22], na.rm = TRUE), 
         Crude.Death.Rate = (Total.Deaths/Population)/100) -> ntrl_deaths

```

## Lets Use our Data

Lets say we were an organization with funding to be able to invest in a country or continents health care. Where should we?

### How do we make this choice

Things to consider

-   Continents population and death
-   Continents crude death rate and how is their death rate trending?

## Lets find what continents have the highest deaths

```{r echo=FALSE}
table_summary <- ntrl_deaths %>% 
  group_by(Continent) %>% 
  summarise(Highest_Death_Count = max(Total.Deaths, na.rm = TRUE)) %>%
  arrange(desc(Highest_Death_Count)) # Sorts highest to lowest

#creates a nice table
table_summary %>%
  mutate(Highest_Death_Count = scales::comma(Highest_Death_Count)) %>%
  kbl(col.names = c("Continent", "Highest Single-Year Total"),
      caption = "Peak Natural Mortality Records by Continent") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = F, 
                position = "left") %>%
  # Header: Navy Blue with White Text
  row_spec(0, background = "midnightblue", color = "white", bold = TRUE) %>%
  # Highlight the Data Column: Light Blue background
  column_spec(2, background = "#F0F7FB", color = "midnightblue", bold = TRUE) %>%
  # Add a border to separate columns
  column_spec(1, border_right = TRUE, bold = TRUE)

#a more better looking bar graph
ntrl_deaths %>% 
  ggplot(aes(x = reorder(Continent, Total.Deaths, FUN = median), 
             y = Total.Deaths, 
             fill = factor(Year))) + # This gives each year a different color
  geom_boxplot(alpha = 0.7, outlier.color = "red") + 
  scale_y_log10(labels = scales::comma) + 
  facet_wrap(~Year) + 
  theme_bw() +
  labs(title = "Natural Deaths by Continent", 
       subtitle = "Color-coded by Year",
       y = "Deaths (Log Scale)", x = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") # Hides the legend since the facet labels already show the year


```

## Continents with the Most Deaths

If we look at the graph, we notice that the continent of Asia has the most death throughout the 4 years.

### Why is this?

Lets look deeper as to why this is.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ntrl_deaths %>% group_by(Continent) %>% 
  summarise(Average.Death.Rate = mean(Crude.Death.Rate))

```

The data here shows that the average death rate in higher in Africa. But how is this when there are morepe total deaths in Asia. To understand this, lets learn about the values

-   what is crude death rate?
-   what does it mean for Asia and their deaths?
-   How can we choose between these 2 continents?

------------------------------------------------------------------------

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ntrl_deaths %>% ggplot(aes(x = Population, 
                           y = Total.Deaths, 
                           color = Continent)) + 
  geom_point() + scale_x_log10() + scale_y_log10() + facet_wrap(~Year)
```

### Based on this graph, what conclusion can we make?

The real reason why Asia has the highest total deaths. This graph says "more people equals more deaths". Whats the easiest conclusion for Asia?

Asia has more people overall, so its a given that theyll have more deaths overall.

## Death Rates Over Time Based in Continents

```{r, echo = FALSE, warning = FALSE, message = FALSE}
ntrl_deaths %>% 
  group_by(Continent, Year) %>% 
  summarise(Year, 
            Average.Death.Rate = mean(Crude.Death.Rate)) -> Avg.DthRt.Continent

Avg.DthRt.Continent %>% ggplot(aes(x = Year, 
                                   y = Average.Death.Rate, 
                                   color = Continent)) + 
                                 geom_line() + labs(title = "Death Rates Throughout the Years", x = "Year", y = "Death rates per 1000")
```

We could choose Asia based on their total deaths, but...?

what about Africa?

Which continent should we actually choose?

## [Insert title]

```{r}
disease_country_count <- death_ds %>%
  pivot_longer(
    cols = -c(Country.Territory, Code, Year, Total.Death),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  filter(Deaths > 0) %>%
  group_by(Disease) %>%
  summarize(
    Countries.Affected = n_distinct(Country.Territory),
    .groups = "drop"
  ) %>%
  arrange(desc(Countries.Affected))

disease_country_count
```

The above information shows which diseases effected the most countries. Below I found which diease effected the most countries and which effected the least.This could help by putting this into anotther study to find which disease should possoibly be researched more to find a possible cure.

```{r}
most_affected <- disease_country_count %>%
  slice_max(Countries.Affected, n = 1)

least_affected <- disease_country_count %>%
  slice_min(Countries.Affected, n = 1)

most_affected
least_affected
```

```{r}
disease_totals <- death_ds %>%
  pivot_longer(
    cols = -c(Country.Territory, Code, Year, Total.Death),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  group_by(Disease) %>%
  summarize(
    Total.Deaths = sum(Deaths, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(Total.Deaths > 0) %>%
  arrange(desc(Total.Deaths))
```

```{r}
disease_totals %>% arrange(desc(Total.Deaths)) %>% slice(1)
disease_totals %>% arrange(Total.Deaths) %>% slice(1)
```

```{r}
#code with all disease over the year from the other graph
#ggplot(combined_disease_data, aes(x = Year, y = Deaths, color = Disease, group = Disease)) +
  #geom_line(linewidth = 1) +
  #geom_point() +
  #scale_y_continuous(labels = label_comma()) +
  #theme_minimal() +
  #labs(title = "Total Deaths by Disease (1990-2015)",
       #x = "Year",
       #y = "Total Global Deaths")


```
