---
title: "Cause_o_Death"
output: slidy_presentation
author: Jorge Alejandro, Mia Alanis, Bryce Borak
---

```{r, echo = FALSE, }
library(tidyverse)
library(dplyr)
library(shiny)
library(ggplot2)
library(scales)
library(kableExtra)
library(sf)
install.packages("rnaturalearth", repos = "https://rstudio.com")
library(rnaturalearth)
library(rnaturalearthdata)
library(stringr)
```


```{r, echo = FALSE}
library(tidyverse)
library(shiny)

#setting up the datasets we will be using for our projects
  death_ds <- read.csv("cause_of_deaths.csv")
  popu_ds <- read.csv("world_population.csv")
  
  #The issue im having now is that ive realized that the years in the population data set are columns and the year in the deaths data set are rows. Im trying to figure out a way where we can have the population info combine with the death data set if the years match up, might have to be a question for the professor)

#Im going to add other columns that include things such as "Total recorded deaths" and a calculation for the crude death rate.
  death_ds %>% mutate(Total.Death = rowSums(death_ds[, -c(1:3)], na.rm = TRUE)) -> death_ds
  
#For a comprehensive dataset with population added to it, I made a join so we can analyse it further.
  dth_pplt_ds <- inner_join(death_ds, popu_ds, join_by(Country.Territory))
```

## What our dataset includes

-   What represents the collumns?
-   What are the rows?

```{r}
glimpse(death_ds)
glimpse(popu_ds)
```

Does mortality rate decrease as the years go by?

What countries would require the most medical funding and research? (this question would require insight on deaths NOT caused by natural causes or "Accident")

## Graphs

Does mortality rate decrease as the years go by?

```{r, echo = FALSE}
death_ds %>% ggplot(aes(x = Year, y = Total.Death, )) + geom_point()
  #For this graph to be a little bit more legible, we either have to change what kind of graph we use, or we solve the first issue of trying to combine population with year and use the color of continent.

```

## What countries would require the most medical funding and research?

```{r,echo = FALSE, warning=FALSE, message=FALSE}
#Mia's Work
#cleaning up / orginizing
 death_ds <- read.csv("cause_of_deaths.csv")
  popu_ds <- read.csv("world_population.csv")

#this is filters to just the years correlated to the population data set
death_ds_filtered <- death_ds %>% 
  filter(Year %in% c(1970, 1980, 1990, 2000, 2010, 2015, 2020, 2022))

combine_data <- inner_join(death_ds_filtered, popu_ds, join_by(Country.Territory))
```

```{r, echo = FALSE, warning=FALSE, message=FALSE}
#Mia's Work

#filtering to only a single year and getting rid of population columns from other years

#may want to delete the ones that we cants use (1970, 1980, 2020, 2022)

#1970 but does not have death info
data_of_1970 <- combine_data %>% 
  filter(Year == "1970") %>% 
  select(-c(X1990.Population, X1980.Population, X2000.Population, X2010.Population, X2015.Population, X2020.Population, X2022.Population))

#1980 but does not have death info
data_of_1980 <- combine_data %>% 
  filter(Year == "1980") %>% 
  select(-c(X1970.Population, X1990.Population, X2000.Population, X2010.Population, X2015.Population, X2020.Population, X2022.Population))

#1990
data_of_1990 <- combine_data %>% 
  filter(Year == "1990") %>% 
  select(-c(X1970.Population, X1980.Population, X2000.Population, X2010.Population, X2015.Population, X2020.Population, X2022.Population))

#2000
data_of_2000 <- combine_data %>% 
  filter(Year == "2000") %>% 
  select(-c(X1970.Population, X1980.Population, X1990.Population, X2010.Population, X2015.Population, X2020.Population, X2022.Population))

#2010
data_of_2010 <- combine_data %>% 
  filter(Year == "2010") %>% 
  select(-c(X1970.Population, X1980.Population, X2000.Population, X1990.Population, X2015.Population, X2020.Population, X2022.Population))

#2015
data_of_2015 <- combine_data %>% 
  filter(Year == "2015") %>% 
  select(-c(X1970.Population, X1980.Population, X2000.Population, X2010.Population, X1990.Population, X2020.Population, X2022.Population))

#2020 but does not have death info
data_of_2020 <- combine_data %>% 
  filter(Year == "2020") %>% 
  select(-c(X1970.Population, X1980.Population, X2000.Population, X2010.Population, X2015.Population, X1990.Population, X2022.Population))
  
#2022 but does not have death info
data_of_2022 <- combine_data %>% 
  filter(Year == "2022") %>% 
  select(-c(X1970.Population, X1980.Population, X1990.Population, X2010.Population, X2015.Population, X2020.Population, X2000.Population))
```

```{r, echo = FALSE, warning=FALSE, message=FALSE}
#Mia's work
#answering the question
#What countries would require the most medical funding and research?
#(this question would require insight on deaths NOT caused by natural causes or "Accident")

#need to filter out deaths that arent natural causes or violent
non_diseases <- c("Drowning", "Interpersonal.Violence", "Drug.Use.Disorders", "Alcohol.Use.Disorders", "Self.harm", "Exposure.to.Forces.of.Nature", "Environmental.Heat.and.Cold.Exposure", "Conflict.and.Terrorism", "Road.Injuries", "Poisonings", "Fire..Heat..and.Hot.Substances", "Maternal.Disorders")

diseases <- death_ds %>% 
  select(-c(non_diseases, Year, Country.Territory, Code)) %>% 
  names()

clean_disease_names <- gsub("\\.", " ", diseases)

#death data with just diseases
disease_deaths_data <- death_ds_filtered %>% 
  select(-c(non_diseases)) 

#this replaces "." in column names to " "
#save for the end because code wont work
#colnames(disease_deaths_data) <- gsub("\\.", " ", colnames(disease_deaths_data))

#death data with just diseases combine with population data
disease_data_pop <- inner_join(disease_deaths_data, popu_ds, join_by(Country.Territory))

#testing dataset with just diseases
new_1990_data <- data_of_1990 %>% select(-non_diseases)
new_2000_data <- data_of_2000 %>% select(-non_diseases)
new_2010_data <- data_of_2010 %>% select(-non_diseases)
new_2015_data <- data_of_2015 %>% select(-non_diseases)
```

```{r, echo = FALSE, warning=FALSE, message=FALSE}

#Should we use this? Maybe to narrow down to which disease to further do research in?
#first graph shows the rise in death of all diseases

#second shows all the diseases and its incline or decline over the years

#third shows top 5 diseases over the years

total_disease_death_1990 <- colSums(new_1990_data[diseases])
total_disease_death_2000 <- colSums(new_2000_data[diseases])
total_disease_death_2010 <- colSums(new_2010_data[diseases])
total_disease_death_2015 <- colSums(new_2015_data[diseases])

# Convert each vector into a small dataframe
df1990 <- data.frame(Disease = names(total_disease_death_1990), Deaths = total_disease_death_1990, Year = 1990)
df2000 <- data.frame(Disease = names(total_disease_death_2000), Deaths = total_disease_death_2000, Year = 2000)
df2010 <- data.frame(Disease = names(total_disease_death_2010), Deaths = total_disease_death_2010, Year = 2010)
df2015 <- data.frame(Disease = names(total_disease_death_2015), Deaths = total_disease_death_2015, Year = 2015)

# Stack them all together
combined_disease_data <- rbind(df1990, df2000, df2010, df2015)
combined_disease_data <- combined_disease_data %>%
  mutate(Disease = str_to_title(gsub("\\.", " ", Disease)))


#Sum up all individual disease totals into one grand total per year
yearly_grand_totals <- data.frame(
  Year = c(1990, 2000, 2010, 2015),
  Total_Deaths = c(
    sum(total_disease_death_1990, na.rm = TRUE),
    sum(total_disease_death_2000, na.rm = TRUE),
    sum(total_disease_death_2010, na.rm = TRUE),
    sum(total_disease_death_2015, na.rm = TRUE)
  )
)
```

## The Total Disease Death Count from 1990 - 2015
Ever wondered how much death occurs just by disease alone? As we know the population is rapidly growing, but is there a growth in disease death as well?

```{r, echo=FALSE, fig.width=10, fig.height=6}
#Graph of Total Global Diseases Through the Years
ggplot(yearly_grand_totals, aes(x = Year, y = Total_Deaths)) +
  geom_line(color = "firebrick", linewidth = 1) +
  geom_point(color = "firebrick", size = 3) +
  # Adds exact amounts as labels above the points
  geom_text(aes(label = comma(Total_Deaths)), vjust = -1.7, size = 3.5) +
  scale_y_continuous(labels = label_comma(), 
                     expand = expansion(mult = c(0.1, 0.2))) + # Adds headroom for labels
  theme_minimal() +
  labs(title = "Total Global Disease Deaths (1990-2015)",
       subtitle = "Aggregated sum of all recorded diseases",
       x = "Year",
       y = "Total Deaths")
```

## All Diseases

What are the Diseases we are referring to?

```{r, echo = FALSE}
clean_disease_names
```

## Comparing Graphs

How has the death amount changed over the years for each specific diseases?

```{r, echo = FALSE, fig.width=8, fig.height=6}
#Sum up total deaths for every disease across all years
disease_totals <- combined_disease_data %>%
  group_by(Disease) %>%
  summarise(Total = sum(Deaths, na.rm = TRUE)) %>%
  arrange(desc(Total)) # Sort highest to lowest

#Just the top 5 names
top_5_names <- head(disease_totals$Disease, 5)

#Filter your main data
top_5_data <- combined_disease_data %>%
  filter(Disease %in% top_5_names)

#Graph of all diseases
ggplot(combined_disease_data, aes(x = Year, y = Deaths, color = Disease, group = Disease)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(labels = label_comma()) +
  theme_minimal() +
  labs(title = "Total Deaths by Disease (1990-2015)",
       x = "Year",
       y = "Total Global Deaths")
```

Lets simplify further more to just the top 5 disease with the highest death counts. 

```{r, echo=FALSE}
#Graph of the top 5 disease
ggplot(top_5_data, aes(x = Year, y = Deaths, color = Disease, group = Disease)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = label_comma()) +
  theme_minimal() +
  labs(title = "Top 5 Deadliest Diseases (1990-2015)",
       subtitle = "Filtered for highest total global death toll",
       x = "Year",
       y = "Total Deaths")

```

In summary, the number one deaths by a disease is Cardiovascular Disease.

```{r, echo = FALSE, fig.width=12, fig.height=8}
#code that orginzes each year long ways
#might not use

#2010
continent_data_2010 <- new_2010_data %>%
  group_by(Continent) %>%
  summarise(across(Meningitis:Acute.Hepatitis, sum, na.rm = TRUE))

long_data_2010 <- continent_data_2010 %>%
  pivot_longer(cols = Meningitis:Acute.Hepatitis, 
               names_to = "Disease", 
               values_to = "Deaths")
#1990
continent_data_1990 <- new_1990_data %>%
  group_by(Continent) %>%
  summarise(across(Meningitis:Acute.Hepatitis, sum, na.rm = TRUE))

long_data_1990 <- continent_data_1990 %>%
  pivot_longer(cols = Meningitis:Acute.Hepatitis, 
               names_to = "Disease", 
               values_to = "Deaths")

#2000
continent_data_2000 <- new_2000_data %>%
  group_by(Continent) %>%
  summarise(across(Meningitis:Acute.Hepatitis, sum, na.rm = TRUE))

long_data_2000 <- continent_data_2000 %>%
  pivot_longer(cols = Meningitis:Acute.Hepatitis, 
               names_to = "Disease", 
               values_to = "Deaths")

#2015
continent_data_2015 <- new_2015_data %>%
  group_by(Continent) %>%
  summarise(across(Meningitis:Acute.Hepatitis, sum, na.rm = TRUE))

long_data_2015 <- continent_data_2015 %>%
  pivot_longer(cols = Meningitis:Acute.Hepatitis, 
               names_to = "Disease", 
               values_to = "Deaths")
```

## If I Were A Reasercher, Where Would I Go?

This graph shows the years: 1990, 2000, 2010, 2015, and comparing the leading cause of death by Continent.

```{r, echo=FALSE, fig.width=12, fig.height=8}

# Combine all years into one master list
all_years_long <- bind_rows(
  long_data_1990 %>% mutate(Year = 1990),
  long_data_2000 %>% mutate(Year = 2000),
  long_data_2010 %>% mutate(Year = 2010),
  long_data_2015 %>% mutate(Year = 2015)
)

# Extract only the top disease for each continent per year
top_diseases_per_continent <- all_years_long %>%
  group_by(Year, Continent) %>%
  slice_max(Deaths, n = 1)
top_diseases_per_continent <- top_diseases_per_continent %>%
  mutate(Disease = str_replace_all(Disease, "\\.", " "))

ggplot(top_diseases_per_continent, aes(x = Continent, y = Deaths, fill = Disease)) +
  geom_col() +
  # Adds the disease name inside or above the bar
  facet_wrap(~Year) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.3))) +
  theme_minimal() +
  labs(title = "Leading Cause of Death by Continent",
       subtitle = "Comparing the #1 disease across years",
       y = "Total Deaths") +
  theme(legend.position = "right") # Disease names are already on the bars

```

As you can see in all four years, Asia stood on top. Also, cardiovascular was the highest disease for every continent in 2010 and 2015.

## Deeper Dive Into Asia

Here is a visual map of Asia's Cardiovascular Deaths, labeling the top 5 Countries.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Get world map data as an 'sf' object
world <- ne_countries(scale = "medium", returnclass = "sf")

# Filter specifically for Asia
asia_map <- world[world$continent == "Asia", ]

map_with_data <- asia_map %>%
  left_join(new_2015_data, by = c("name" = "Country.Territory"))

#the top 5 countries
top_5_countries <- map_with_data %>%
  arrange(desc(Cardiovascular.Diseases)) %>%
  head(5)
  
top_5_countries <- top_5_countries %>%
  mutate(
    my_nudge_y = case_when(
      name == "Japan"     ~ -6.0,  # Moves Japan down by 2 degrees
      name == "Indonesia" ~ -2.0,  # Moves Indonesia down by 3.5 degrees
      TRUE                ~ 0.0    # Everyone else stays at 0
    )
  )

# Plotting deaths (e.g., Cardiovascular)
ggplot(data = map_with_data) +
  geom_sf(aes(fill = Cardiovascular.Diseases), color = "black", size = 0.1) +
  geom_sf_text(data = top_5_countries, aes(label = name), 
                 nudge_y = top_5_countries$my_nudge_y,
               size = 3, fontface = "bold", nudge_y = 0) +
  scale_fill_distiller(
    palette = "YlOrRd", 
    direction = 1, 
    labels = scales::comma,
    na.value = "grey90"
  ) + # Added the missing + here
  theme_void() + 
  labs(fill = "Deaths", title = "Cardiovascular Deaths Across Asia")
```

Each country's Cardiovascular Deaths in Asia, sorted from the highest to lowest.

```{r, echo=FALSE, fig.width=8, fig.height=15}
new_2015_data %>% 
  filter(Continent == "Asia") %>% 
  group_by(Country.Territory) %>% 
  summarize(Total_CVD = sum(Cardiovascular.Diseases, na.rm = TRUE)) %>% 
  arrange(desc(Total_CVD)) %>% 
  kbl(col.names = c("Country", "Total Cardiovascular Deaths")) %>% # Set clean names here
  kable_paper() %>% 
  scroll_box(height = "500px", width = "1000px")
```
