---
title: "Cause_of_Death"
author: "Jorge Alejandro, Mia Alanis, Bryce Borak"
output: slidy_presentation
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(shiny)

#setting up the datasets we will be using for our projects
  death_ds <- read.csv("cause_of_deaths.csv")
  popu_ds <- read.csv("world_population.csv")

#Im going to add other columns that include things such as "Total recorded deaths" and a calculation for the crude death rate.
  death_ds %>% mutate(Total.Death = rowSums(death_ds[, -c(1:3)], na.rm = TRUE)) -> death_ds
  
  death_ds %>% 
    filter(Year %in% c(1970, 1980, 1990, 2000, 2010, 2015, 2020, 2022)) -> death_ds
  
#This changes the population dataset to a viable format that we could use to join the death and the population datasheets
  popu_ds %>% 
    pivot_longer(
      cols = matches("^X\\d"),
      names_to = "Year",
      values_to = "Population"
    ) %>% 
    mutate(Year = as.numeric(gsub("X|\\.Population", "", Year))) -> popu_longer
  
#For a comprehensive dataset with population added to it, I made a join so we can analyse it further.
  dth_pplt_ds <- inner_join(death_ds, popu_longer, join_by(Country.Territory, Year))
  
#We might also want to use the numbers of the death rate of a population so im going to mutate the dataset to have that in the final column
  
  dth_pplt_ds %>% mutate(Crude.Death.Rate = (Total.Death / Population) * 1000, .after = "Total.Death") -> dth_pplt_ds
  
non_diseases <- c("Drowning", "Interpersonal.Violence", "Drug.Use.Disorders", "Alcohol.Use.Disorders", "Self.harm", "Exposure.to.Forces.of.Nature", "Environmental.Heat.and.Cold.Exposure", "Conflict.and.Terrorism", "Road.Injuries", "Poisonings", "Fire..Heat..and.Hot.Substances", "Maternal.Disorders")
```
## Our Data
Our data come from Kaggle. "Cause of Deaths" and "World Population", both CSV data sets.

## What is Shown in our Data Set
```{r}
tibble(dth_pplt_ds)
```

## What Makes up the Majority of Our Columns?
### Our dataset show various different death causes, both natural and unnatural.
```{r, echo = FALSE, warning = FALSE}
print(colnames(dth_pplt_ds)[5:34])
```

## What Can we Learn with Our Data
Our main focus is the "Cause of Deaths Data set", So where does the "World Population" data set come into play?

How are we using both these data set and what can we learn?

- Continents and countries
- Population and how we calculate "Death rate"
- As the years go by, population increases. Does this affect the death rate significantly?
- Which diesease effects the most countries 

## Continents with the Most Deaths
```{r, echo=FALSE, message=FALSE, warning=FALSE}

dth_pplt_ds %>% 
  select(-c(non_diseases, Total.Death, Crude.Death.Rate)) -> ntrl_deaths

ntrl_deaths %>% 
  mutate(Total.Deaths = rowSums(ntrl_deaths[, 4:22], na.rm = TRUE), 
         Crude.Death.Rate = (Total.Deaths/Population)*1000) -> ntrl_deaths

#Now we can try to find what continent has the highest number of deaths
ntrl_deaths %>% group_by(Continent) %>% 
  summarise(Highest_Death_Continent = max(Total.Deaths))

ntrl_deaths %>% ggplot(aes(x = Continent, 
                           y = Total.Deaths, 
                           color = Continent)) + geom_boxplot() + 
  scale_y_log10() + 
  facet_wrap(~Year)
```

## Continents with the Most Deaths

  If we look at the graph, we notice that the continent of Asia has the most death throughout the 4 years.
  
### Why is this?

Lets look deeper as to why this is.
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ntrl_deaths %>% group_by(Continent) %>% 
  filter(Year == 2000) %>%
  summarise(Average.Death.Rate = mean(Crude.Death.Rate))

```
  The data here shows that the average death rate in higher in Africa. But how is this when more people die in total in Asia. To understand this, lets learn about the values
  
- what is crude death rate?
-what does it mean for asia and their deaths?
-How can we choose between these 2 continents?

---
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ntrl_deaths %>% ggplot(aes(x = Population, 
                           y = Total.Deaths, 
                           color = Continent)) + 
  geom_line() + facet_wrap(~Year)
```
Based on this graph, what conclusion can we make?



---
```{r}
disease_country_count <- death_ds %>%
  pivot_longer(
    cols = -c(Country.Territory, Code, Year, Total.Death),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  filter(Deaths > 0) %>%
  group_by(Disease) %>%
  summarize(
    Countries.Affected = n_distinct(Country.Territory),
    .groups = "drop"
  ) %>%
  arrange(desc(Countries.Affected))

disease_country_count
```
The above information shows which diseases effected the most countries. Below I found which diease effected the most countries and which effected the least.This could help by putting this into anotther study to find which disease should possoibly be researched more to find a possible cure. 
```{r}
most_affected <- disease_country_count %>%
  slice_max(Countries.Affected, n = 1)

least_affected <- disease_country_count %>%
  slice_min(Countries.Affected, n = 1)

most_affected
least_affected
```
```{r}
disease_totals <- death_ds %>%
  pivot_longer(
    cols = -c(Country.Territory, Code, Year, Total.Death),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  group_by(Disease) %>%
  summarize(
    Total.Deaths = sum(Deaths, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(Total.Deaths > 0) %>%
  arrange(desc(Total.Deaths))
```


```{r}
disease_totals %>% arrange(desc(Total.Deaths)) %>% slice(1)
disease_totals %>% arrange(Total.Deaths) %>% slice(1)
```

