---
title: "Cause_of_Death"
author: "Jorge Alejandro, Mia Alanis, Bryce Borak"
output: slidy_presentation
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(shiny)
library(knitr)
library(ggplot2)
library(kableExtra)
library(stringr)
library(patchwork)

#setting up the datasets we will be using for our projects
  death_ds <- read.csv("cause_of_deaths.csv")
  popu_ds <- read.csv("world_population.csv")

#Im going to add other columns that include things such as "Total recorded deaths" and a calculation for the crude death rate.
  death_ds %>% mutate(Total.Death = rowSums(death_ds[, -c(1:3)], na.rm = TRUE)) -> death_ds
  
  death_ds %>% 
    filter(Year %in% c(1970, 1980, 1990, 2000, 2010, 2015, 2020, 2022)) -> death_ds
  
#This changes the population dataset to a viable format that we could use to join the death and the population datasheets
  popu_ds %>% 
    pivot_longer(
      cols = matches("^X\\d"),
      names_to = "Year",
      values_to = "Population"
    ) %>% 
    mutate(Year = as.numeric(gsub("X|\\.Population", "", Year))) -> popu_longer
  
#For a comprehensive dataset with population added to it, I made a join so we can analyse it further.
  dth_pplt_ds <- inner_join(death_ds, popu_longer, join_by(Country.Territory, Year))
  
#We might also want to use the numbers of the death rate of a population so im going to mutate the dataset to have that in the final column
  
  dth_pplt_ds %>% mutate(Crude.Death.Rate = (Total.Death / Population) * 100, .after = "Total.Death") -> dth_pplt_ds

```

## Our Data

::: {style="float: left; width: 45%; border-right: 2px solid steelblue; padding-right: 20%;"}
### **Raw Input**

**World Population** \* 234 Rows \| 17 Columns \* [Source Link](https://www.kaggle.com/datasets/iamsouravbanerjee/world-population-dataset)

**Causes of Death** \* 6,120 Rows \| 34 Columns \* [Source Link](https://www.kaggle.com/datasets/iamsouravbanerjee/cause-of-deaths-around-the-world)
:::

::: {style="float: right; width: 45%; color: midnightblue;"}
### **Final Data Set**

-   **796** Rows
-   **45** Columns

**Key Operations:** 1. **Filter:** Decadal years (1970-2022) 2. **Pivot:** Population into long format 3. **Join:** Inner join by Country/Year 4. **Mutate:** Added Crude Death Rate
:::

```{r, echo=FALSE, fig.width=8, fig.height=15}
dth_pplt_ds %>%
  head(100) %>%
  # Remove the comma from Year and format numbers
  mutate(Year = as.character(Year),
         across(where(is.numeric), ~scales::comma(.))) %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("bordered", "condensed", "hover"), 
    full_width = FALSE,  # This keeps the table from stretching
    position = "center",  # Centers the smaller table on the page
    font_size = 18    # Manually shrink the text size (default is usually 16)
  ) %>%
  row_spec(0, background = "steelblue", color = "white") %>%
  # Explicitly set column widths to keep it "thin"
  column_spec(1:ncol(dth_pplt_ds), border_left = TRUE, border_right = TRUE) %>% 
  scroll_box(width = "1400px", height = "350px") # Set a fixed pixel width instead of 100%
```

## What is Shown in our Data Set

### What Makes up the Majority of Our Columns?

Our dataset show various different death causes, both natural and unnatural.

```{r, echo = FALSE, warning = FALSE, fig.width=8, fig.height=13 }
all_cause_cols <- colnames(dth_pplt_ds)[5:34]

non_diseases <- c("Drowning", "Interpersonal.Violence", "Drug.Use.Disorders", "Alcohol.Use.Disorders", "Self.harm", "Exposure.to.Forces.of.Nature", "Environmental.Heat.and.Cold.Exposure", "Conflict.and.Terrorism", "Road.Injuries", "Poisonings", "Fire..Heat..and.Hot.Substances", "Maternal.Disorders")

clean_non_diseases <- str_replace_all(non_diseases, "\\.", " ")

disease_cols <- setdiff(all_cause_cols, non_diseases)

clean_diseases_cols <- str_replace_all(disease_cols, "\\.", " ")

#this is just separating the columns with out the death types
other_columns <- setdiff(colnames(dth_pplt_ds), all_cause_cols)

clean_other_cols <- str_replace_all(other_columns, "\\.", " ")

other_cols_table <- data.frame(
  Other_Columns = clean_other_cols
)

max_ln_all <- max(length(clean_other_cols), 
                  length(clean_diseases_cols), 
                  length(clean_non_diseases))

#this is creating the combined data frame
combined_dictionary <- data.frame(
  Other_Columns = c(clean_other_cols, rep("", max_ln_all - length(clean_other_cols))),
  Disease_Causes = c(clean_diseases_cols, rep("", max_ln_all - length(clean_diseases_cols))),
  Non_Disease_Causes = c(clean_non_diseases, rep("", max_ln_all - length(clean_non_diseases)))
)

#this is to create a table to show both diseases and non-diseases but also the other columns
max_ln <- max(length(clean_diseases_cols) , length(clean_non_diseases))

comparison_table <- data.frame(
  Disease_Causes = c(clean_diseases_cols, rep(NA, max_ln - length(clean_diseases_cols))),
  Non_Disease_Causes = c(clean_non_diseases, rep(NA, max_ln - length(clean_non_diseases)))
)

comparison_table[is.na(comparison_table)] <- ""

#this create a better looking Table
combined_dictionary %>%
  kbl(col.names = c("Data Columns", "Diseases Columns", "Non-Disease Columns"),
      caption = "Dataset Column Classification",
      align = "lll") %>%
  kable_styling(bootstrap_options = c("bordered", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 26) %>%
  # Header Styling
  row_spec(0, background = "steelblue", color = "white", bold = TRUE) %>%
  # Column Boxing and Widths
  column_spec(1, border_left = TRUE, border_right = TRUE, width = "15em", bold = TRUE) %>%
  column_spec(2, border_right = TRUE, width = "18em") %>%
  column_spec(3, border_right = TRUE, width = "18em") %>% 
  scroll_box(width = "1400px", height = "550px")
```

------------------------------------------------------------------------

::: {style="float: left; width: 48%; padding-top: 10px;"}
### **The Data Collaboration**

The **"Cause of Deaths"** dataset provides the *counts*, but the **"World Population"** dataset provides the *context*.

By merging them, we move from raw totals to meaningful **rates**. This allows us to answer not just *"how many died?"*, but **"how heavily was the population actually impacted?"**

<br>

### **The Investment Perspective**

**Imagine we are a global healthcare fund.** How do we choose where to invest? 

We don't just look at who has the most deaths; we analyze:
- **The Crude Death Rate:** The true burden on the people.
- **The Trend:** Is the crisis improving or worsening?
:::

::: {style="float: right; width: 45%; background-color: #F0F8FF; padding: 10px; border-radius: 12px; border: 2px solid #4682B4; color: #191970;"}
### **Our Core Questions**

- **Regional Burden:** Which continents face the heaviest impact from biological diseases?


- **Fair Comparison:** How do we level the playing field between small nations and global giants?


- **Trends Over Time:** How has the relationship between population and mortality evolved?


- **Global Reach:** Which specific diseases are the most widespread across international borders?
:::


```{r echo=FALSE, warning = FALSE}
## Continents with the Most Deaths

ntrl_deaths <- dth_pplt_ds %>% 
  select(-c(all_of(non_diseases), Total.Death, Crude.Death.Rate))

ntrl_deaths <- ntrl_deaths %>% 
  mutate(
    Total.Deaths = rowSums(across(all_of(disease_cols)), na.rm = TRUE), 
    # Standard crude rate is usually per 1,000 or 100,000
    # Your current (Total/Pop)/100 is essentially a percentage.
    Crude.Death.Rate = (Total.Deaths / Population) * 100000
  )
```

---

```{r echo=FALSE}
table_summary <- ntrl_deaths %>% 
  group_by(Continent) %>% 
  summarise(Highest_Death_Count = max(Total.Deaths, na.rm = TRUE)) %>%
  arrange(desc(Highest_Death_Count)) # Sorts highest to lowest

#creates a nice table
table_summary %>%
  mutate(Highest_Death_Count = scales::comma(Highest_Death_Count)) %>%
  kbl(col.names = c("Continent", "Highest Single-Year Total"),
      caption = "Peak Natural Mortality Records by Continent") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = F, 
                position = "left") %>%
  # Header: Navy Blue with White Text
  row_spec(0, background = "midnightblue", color = "white", bold = TRUE) %>%
  # Highlight the Data Column: Light Blue background
  column_spec(2, background = "#F0F7FB", color = "midnightblue", bold = TRUE) %>%
  # Add a border to separate columns
  column_spec(1, border_right = TRUE, bold = TRUE)

#a more better looking bar graph
ntrl_deaths %>% 
  ggplot(aes(x = reorder(Continent, Total.Deaths, FUN = median), 
             y = Total.Deaths, 
             fill = factor(Year))) + # This gives each year a different color
  geom_boxplot(alpha = 0.7, outlier.color = "red") + 
  scale_y_log10(labels = scales::comma) + 
  facet_wrap(~Year) + 
  theme_bw() +
  labs(title = "Natural Deaths by Continent", 
       subtitle = "Color-coded by Year",
       y = "Deaths (Log Scale)", x = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") # Hides the legend since the facet labels already show the year


```


### Continents with the Most Deaths

If we look at the graph, we notice that the continent of Asia has the most death throughout the 4 years.

### Why is this?

Lets look deeper as to why this is.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ntrl_deaths %>% group_by(Continent) %>% 
  summarise(Average.Death.Rate = mean(Crude.Death.Rate))

```

The data here shows that the average death rate in higher in Africa. But how is this when there are morepe total deaths in Asia. To understand this, lets learn about the values

-   what is crude death rate?
-   what does it mean for Asia and their deaths?
-   How can we choose between these 2 continents?

---

## Fair Comparison

### How do we level the playing field between small nations and global giants?

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ntrl_deaths %>% ggplot(aes(x = Population, 
                           y = Total.Deaths, 
                           color = Continent)) + 
  geom_point() + scale_x_log10() + scale_y_log10() + facet_wrap(~Year)
```

### Based on this graph, what conclusion can we make?

The real reason why Asia has the highest total deaths. This graph says "more people equals more deaths". Whats the easiest conclusion for Asia?

Asia has more people overall, so its a given that theyll have more deaths overall.

### Death Rates Over Time Based in Continents

```{r, echo = FALSE, warning = FALSE, message = FALSE}
ntrl_deaths %>% 
  group_by(Continent, Year) %>% 
  summarise(Year, 
            Average.Death.Rate = mean(Crude.Death.Rate)) -> Avg.DthRt.Continent

Avg.DthRt.Continent %>% ggplot(aes(x = Year, 
                                   y = Average.Death.Rate, 
                                   color = Continent)) + 
                                 geom_line() + labs(title = "Death Rates Throughout the Years", x = "Year", y = "Death rates per 1000")
```

We could choose Asia based on their total deaths, but...?

what about Africa?

Which continent should we actually choose?

---

## Regional Burden

### Which continents face the heaviest impact from biological diseases?

```{r, echo=FALSE}
ntrl_deaths <- ntrl_deaths %>% 
  select(-c(Code, Rank, CCA3, Capital, Density..per.km.., Growth.Rate, World.Population.Percentage, Area..km..))  %>% 
  relocate(Continent, .after = Country.Territory) %>% 
  relocate(Total.Deaths, .after = Acute.Hepatitis)

#Summarize with the Weighted Rate
# We save this to a new object for your plot/table
dis_crude_rate <- ntrl_deaths %>% 
  group_by(Continent, Year) %>%
  summarize(
    # We must SUM the deaths and population for the continent
    Total_Deaths_Continent = sum(Total.Deaths, na.rm = TRUE),
    # Calculate the rate here so the next line can 'arrange' by it
    Crude.Death.Rate = (Total_Deaths_Continent / sum(Population, na.rm = TRUE)) * 100000,
    .groups = "drop"
  ) %>%
  # Now arrange works because Crude.Death.Rate exists in this table
  arrange(Year, desc(Crude.Death.Rate))

#disease crude death rate
dis_crude_rate <- ntrl_deaths %>%
  group_by(Continent, Year) %>%
  summarize(
    # name it clearly to distinguish it from the overall Crude Death Rate
    Avg_Dis_Death_Rate = mean((Total.Deaths / Population) * 100000, na.rm = TRUE)
  , .groups = "drop" ) %>%
  arrange(desc(Avg_Dis_Death_Rate))

```

::: {.columns}

::: {.column width="65%"}
```{r, echo=FALSE}
# --- STEP 2: GRAPH RENDER ---
ggplot(dis_crude_rate, aes(x = Year, y = Avg_Dis_Death_Rate, color = Continent)) +
  geom_line(linewidth = 2.5) +
  geom_point(size = 4) +
  # Highlighting Europe with a specific color if desired
  scale_color_brewer(palette = "RdYlBu") +
  theme_minimal() +
  labs(title = "The Mortality Trajectory",
       subtitle = "Disease-driven death rates per 100k (1990 - 2015)",
       x = "Year",
       y = "Deaths per 100,000") +
  theme(
    plot.title = element_text(face = "bold", color = "#191970", size = 20),
    legend.position = "bottom",
    # MAKE AXIS WORDING BIGGER
    # Axis Titles ("Year" and "Deaths per 100,000")
    axis.title = element_text(size = 16, face = "bold"), 
    
    # 2. Axis Text (The actual numbers/years on the scales)
    axis.text = element_text(size = 14, color = "black"),
    
    # 3. Legend Text (The continent names)
    legend.text = element_text(size = 12)
  )
```
:::

::: {.column width="35%"}
<style>
.table {
    margin-top: 30px; /* Moves table down from the header */
    border: 2px solid #4682B4; /* Adds a Steelblue border around the whole table */
}
</style>
```{r, echo=FALSE}
# --- STEP 3: TABLE RENDER ---
ntrl_deaths %>%
  filter(Year %in% c(1990, 2015)) %>%
  group_by(Continent, Year) %>%
  summarize(
    Avg_Rate = (sum(Total.Deaths, na.rm = TRUE) / sum(Population, na.rm = TRUE)) * 100000,
    .groups = "drop"
  ) %>%
  # This "spreads" the years into their own columns
  pivot_wider(names_from = Year, values_from = Avg_Rate) %>%
  mutate(Growth = `2015` - `1990`) %>% # Optional: Shows the change
  arrange(desc(`2015`)) %>%
  kable(digits = 1, col.names = c("Continent", "1990 Rate", "2015 Rate", "Absolute Change")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 18) %>%
  # Highlight the 2015 column to show it's the "Current" status
  column_spec(3, bold = TRUE, background = "#F0F8FF") %>%
  row_spec(0, background = "steelblue", color = "white")

```
<div style="background-color: #ffffff; padding: 15px; border-left: 5px solid #191970;color: #191970; font-size: 0.9em; margin-top: 15px; box-shadow: 2px 2px 5px rgba(0,0,0,0.1);">
We show deaths per 100,000 people so that large countries don't overshadow smaller ones. This reveals when major outbreaks or successful medical breakthroughs shifted the numbers.
</div>
:::

:::

---

## Global Reach {data-background="#F0F8FF"}

### Which specific diseases are the most widespread across international borders?

```{r, echo=FALSE}
# Calculate global mean for each disease
global_reach <- dth_pplt_ds %>%
  # We sum all deaths for a disease globally, then divide by total global population
  summarize(across(all_of(disease_cols), 
                   ~ (sum(. , na.rm = TRUE) / sum(Population, na.rm = TRUE)) * 10000)) %>%
  pivot_longer(cols = everything(), 
               names_to = "Disease", 
               values_to = "Global_Avg_Rate") %>%
  arrange(desc(Global_Avg_Rate)) %>%
  slice_head(n = 6)

# Prepare data for the Dot Plot
top_diseases <- global_reach$Disease

dot_plot_data <- dth_pplt_ds %>%
  select(Continent, all_of(top_diseases), Population) %>%
  pivot_longer(cols = all_of(top_diseases), names_to = "Disease", values_to = "Deaths") %>%
  group_by(Continent, Disease) %>%
  # Weighted calculation: Total deaths in continent / Total population in continent
  summarize(Avg_Rate = (sum(Deaths, na.rm = TRUE) / sum(Population, na.rm = TRUE)) * 10000, 
            .groups = "drop") %>%
  mutate(Disease = gsub("\\.", " ", Disease))
```
::: {.columns}

::: {.column width="40%"}
```{r, echo=FALSE}
# Re-using your global_reach processing from before
global_reach %>%
  mutate(Disease = gsub("\\.", " ", Disease)) %>%
  kable(col.names = c("Disease", "Avg. Rate (100k)"), digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 16) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4682B4")
```
<div style="background-color: #ffffff; padding: 15px; border-left: 5px solid #4682B4; color: #191970; font-size: 0.9em; box-shadow: 2px 2px 5px rgba(0,0,0,0.05);">
<b>The "Intensity Gap":</b> The grey lines represent the disparity between regions. A longer line indicates a disease that is highly sensitive to local infrastructure and <b>resource allocation.</b>
</div>
:::

::: {.column width="60%"}
```{r, echo=FALSE}
# Render Dot Plot with high-distinction colors
ggplot(dot_plot_data, aes(x = Avg_Rate, y = reorder(Disease, Avg_Rate), color = Continent)) +
  geom_line(aes(group = Disease), color = "grey85", linewidth = 1.2) +
  geom_point(size = 5, alpha = 0.9) +
  # Using 'Set2' for much better distinction between 6 continents
  scale_color_brewer(palette = "RdYlBu") +
  theme_minimal() +
  labs(title = "The Global Health Gap",
       subtitle = "Average deaths per 100,000 people",
       x = "Deaths per 100k", y = NULL) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", color = "#191970", size = 18),
    axis.text = element_text(size = 11, color = "#333333"),
    legend.text = element_text(size = 10)
  )
```
:::

:::

---

## Lets Focus in on Europe

```{r, echo = FALSE}

europe_ntrl_dths <- ntrl_deaths %>%
  filter(Continent == "Europe") %>%
  pivot_longer(
    cols = -c(Country.Territory, Continent, Year, Total.Deaths, Population),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  filter(Deaths > 0) %>%
  group_by(Disease) %>%
  summarize(
    Countries.Affected = n_distinct(Country.Territory),
    .groups = "drop"
  ) %>%
  arrange(desc(Countries.Affected))

europe_ntrl_dths %>% 
  kable(col.names = c("Causes of Deaths", "Countries Affected by Deaths"), digits = 1) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 16) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4682B4")

```

The above information shows which diseases effected the most countries in Europe. We should further investigate to see the trends of these diseases.

```{r, echo = FALSE, message = FALSE, warning = FALSE}

europe_ntrl_dths %>% filter(Countries.Affected > 40) -> max_europe_ntrl_dths

max_europe_ntrl_dths[[1]] -> disease_names

ntrl_deaths %>% select(all_of(disease_names), Year, Continent) %>% filter(Continent == "Europe") -> for_graph

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
disease_totals <- death_ds %>%
  pivot_longer(
    cols = -c(Country.Territory, Code, Year, Total.Death),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  group_by(Disease) %>%
  summarize(
    Total.Deaths = sum(Deaths, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(Total.Deaths > 0) %>%
  arrange(desc(Total.Deaths))
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
disease_totals %>% arrange(desc(Total.Deaths)) %>% slice(1)
disease_totals %>% arrange(Total.Deaths) %>% slice(1)
```
---

=======
## Trends Over Time

### How has the relationship between population and mortality evolved?

---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
regional_summary <- ntrl_deaths %>%
  filter(Year %in% c(1990, 2000, 2010, 2015)) %>%
  pivot_longer(
    cols = all_of(disease_cols),
    names_to = "Disease",
    values_to = "Death.Rate"
  ) %>%
  group_by(Continent, Disease) %>%
  summarise(
    Avg_Rate = mean(Death.Rate, na.rm = TRUE),
    Max_Rate = max(Death.Rate, na.rm = TRUE),
    Min_Rate = min(Death.Rate, na.rm = TRUE),
    Regional_Gap = Max_Rate - Min_Rate,
    .groups = "drop"
  )
```

```{r}
regional_summary %>%
  group_by(Continent) %>%
  summarise(Avg_Regional_Burden = mean(Avg_Rate, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = reorder(Continent, Avg_Regional_Burden), y = Avg_Regional_Burden, fill = Continent)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Average Disease Burden by Region",
       x = "Continent",
       y = "Average Death Rate") +
  guides(fill = "none")
```

```{r}
ntrl_deaths %>%
  filter(Year %in% c(1990, 2000, 2010, 2015)) %>%
  pivot_longer(
    cols = all_of(disease_cols),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  filter(Deaths > 100) %>%  # removes tiny irrelevant values
  group_by(Disease) %>%
  summarise(Countries.Affected = n_distinct(Country.Territory), .groups = "drop") %>%
  arrange(desc(Countries.Affected)) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(Disease, Countries.Affected), y = Countries.Affected)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = Countries.Affected), hjust = -0.2, size = 5) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Most Widespread Diseases (Global Reach)",
       subtitle = "Top 8 diseases by number of countries affected",
       x = "Disease",
       y = "Countries Affected") +
  ylim(0, NA)
```

