---
title: "Cause_o_Death"
output: slidy_presentation
author: Jorge Alejandro, Mia Alanis, Bryce Borak
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(shiny)

#setting up the datasets we will be using for our projects
  death_ds <- read.csv("cause_of_deaths.csv")
  popu_ds <- read.csv("world_population.csv")
  
  #The issue im having now is that ive realized that the years in the population data set are columns and the year in the deaths data set are rows. Im trying to figure out a way where we can have the population info combine with the death data set if the years match up, might have to be a question for the professor)

#Im going to add other columns that include things such as "Total recorded deaths" and a calculation for the crude death rate.
  death_ds %>% mutate(Total.Death = rowSums(death_ds[, -c(1:3)], na.rm = TRUE)) -> death_ds
  
  death_ds %>% 
    filter(Year %in% c(1970, 1980, 1990, 2000, 2010, 2015, 2020, 2022)) -> death_ds
  
#This changes the population dataset to a viable format that we could use to join the death and the population datasheets
  popu_ds %>% 
    pivot_longer(
      cols = matches("^X\\d"),
      names_to = "Year",
      values_to = "Population"
    ) %>% 
    mutate(Year = as.numeric(gsub("X|\\.Population", "", Year))) -> popu_longer
  
#For a comprehensive dataset with population added to it, I made a join so we can analyse it further.
  dth_pplt_ds <- inner_join(death_ds, popu_longer, join_by(Country.Territory, Year))
  
#We might also want to use the numbers of the death rate of a population so im going to mutate the dataset to have that in the final column
  
  dth_pplt_ds %>% mutate(Crude.Death.Rate = (Total.Death / Population) * 1000, .after = "Total.Death") -> dth_pplt_ds
```
## Our Data
Our data come from Kaggle. "Cause of Deaths" and "World Population", both CSV data sets.

Whyd we choose this data and what are we hoping to learn from it?

## What is Shown in our Data Set
```{r}
tibble(dth_pplt_ds)
```

## What Makes up the Majority of Our Columns?
### Our dataset show various different death causes, both natural and unnatural.
```{r, echo = FALSE, warning = FALSE}
print(colnames(dth_pplt_ds)[5:34])
```
## What Can we Learn with Our Data
Our main focus is the "Cause of Deaths Data set", So where does the "World Population" data set come into play?

How are we using both these data set and what can we learn?

- Continents and countries
- Population and how we calculate "Death rate"
- As the years go by, population increases. DOes this affect the death rate significantly?

## 
```{r, echo = FALSE, warning = FALSE, message = FALSE}
#Lets filter our data further.
non_diseases <- c("Drowning", "Interpersonal.Violence", "Drug.Use.Disorders", "Alcohol.Use.Disorders", "Self.harm", "Exposure.to.Forces.of.Nature", "Environmental.Heat.and.Cold.Exposure", "Conflict.and.Terrorism", "Road.Injuries", "Poisonings", "Fire..Heat..and.Hot.Substances", "Maternal.Disorders")

dth_pplt_ds %>% 
  select(-c(non_diseases, Total.Death, Crude.Death.Rate)) -> ntrl_deaths

ntrl_deaths %>% 
  mutate(Total.Deaths = rowSums(ntrl_deaths[, 4:22], na.rm = TRUE), 
         Crude.Death.Rate = (Total.Deaths/Population)/100) -> ntrl_deaths

#Now we can try to find what continent has the highest number of deaths
ntrl_deaths %>% group_by(Continent) %>% 
  summarise(Highest_Death_Continent = max(Total.Deaths))

ntrl_deaths %>% ggplot(aes(x = Continent, y = Total.Deaths)) + geom_boxplot() + scale_y_log10() + facet_wrap(~Year)
```








```{r, echo = FALSE}

```

## What countries would require the most medical funding and research?


