---
title: "Cause_of_Death"
author: "Jorge Alejandro, Mia Alanis, Bryce Borak"
output: slidy_presentation
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(shiny)
library(knitr)
library(ggplot2)
library(kableExtra)
library(stringr)
library(patchwork)

#setting up the datasets we will be using for our projects
  death_ds <- read.csv("cause_of_deaths.csv")
  popu_ds <- read.csv("world_population.csv")

#Im going to add other columns that include things such as "Total recorded deaths" and a calculation for the crude death rate.
  death_ds %>% mutate(Total.Death = rowSums(death_ds[, -c(1:3)], na.rm = TRUE)) -> death_ds
  
  death_ds %>% 
    filter(Year %in% c(1970, 1980, 1990, 2000, 2010, 2015, 2020, 2022)) -> death_ds
  
#This changes the population dataset to a viable format that we could use to join the death and the population datasheets
  popu_ds %>% 
    pivot_longer(
      cols = matches("^X\\d"),
      names_to = "Year",
      values_to = "Population"
    ) %>% 
    mutate(Year = as.numeric(gsub("X|\\.Population", "", Year))) -> popu_longer
  
#For a comprehensive dataset with population added to it, I made a join so we can analyse it further.
  dth_pplt_ds <- inner_join(death_ds, popu_longer, join_by(Country.Territory, Year))
  
#We might also want to use the numbers of the death rate of a population so im going to mutate the dataset to have that in the final column
  
  dth_pplt_ds %>% mutate(Crude.Death.Rate = (Total.Death / Population) * 1000, .after = "Total.Death") -> dth_pplt_ds

```

## Our Data

::: {style="float: left; width: 45%; border-right: 2px solid steelblue; padding-right: 20%;"}
### **Raw Input**

**World Population** \* 234 Rows \| 17 Columns \* [Source Link](https://www.kaggle.com/datasets/iamsouravbanerjee/world-population-dataset)

**Causes of Death** \* 6,120 Rows \| 34 Columns \* [Source Link](https://www.kaggle.com/datasets/iamsouravbanerjee/cause-of-deaths-around-the-world)
:::

::: {style="float: right; width: 45%; color: midnightblue;"}
### **Final Data Set**

-   **796** Rows
-   **45** Columns

**Key Operations:** 1. **Filter:** Decadal years (1970-2022) 2. **Pivot:** Population into long format 3. **Join:** Inner join by Country/Year 4. **Mutate:** Added Crude Death Rate
:::

```{r, echo=FALSE, fig.width=8, fig.height=15}
dth_pplt_ds %>%
  head(100) %>%
  # Remove the comma from Year and format numbers
  mutate(Year = as.character(Year),
         across(where(is.numeric), ~scales::comma(.))) %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("bordered", "condensed", "hover"), 
    full_width = FALSE,  # This keeps the table from stretching
    position = "center",  # Centers the smaller table on the page
    font_size = 18    # Manually shrink the text size (default is usually 16)
  ) %>%
  row_spec(0, background = "steelblue", color = "white") %>%
  # Explicitly set column widths to keep it "thin"
  column_spec(1:ncol(dth_pplt_ds), border_left = TRUE, border_right = TRUE) %>% 
  scroll_box(width = "1400px", height = "350px") # Set a fixed pixel width instead of 100%
```

## What is Shown in our Data Set

### What Makes up the Majority of Our Columns?

Our dataset show various different death causes, both natural and unnatural.

```{r, echo = FALSE, warning = FALSE, fig.width=8, fig.height=13 }
all_cause_cols <- colnames(dth_pplt_ds)[5:34]

non_diseases <- c("Drowning", "Interpersonal.Violence", "Drug.Use.Disorders", "Alcohol.Use.Disorders", "Self.harm", "Exposure.to.Forces.of.Nature", "Environmental.Heat.and.Cold.Exposure", "Conflict.and.Terrorism", "Road.Injuries", "Poisonings", "Fire..Heat..and.Hot.Substances", "Maternal.Disorders")

clean_non_diseases <- str_replace_all(non_diseases, "\\.", " ")

disease_cols <- setdiff(all_cause_cols, non_diseases)

clean_diseases_cols <- str_replace_all(disease_cols, "\\.", " ")

#this is just separating the columns with out the death types
other_columns <- setdiff(colnames(dth_pplt_ds), all_cause_cols)

clean_other_cols <- str_replace_all(other_columns, "\\.", " ")

other_cols_table <- data.frame(
  Other_Columns = clean_other_cols
)

max_ln_all <- max(length(clean_other_cols), 
                  length(clean_diseases_cols), 
                  length(clean_non_diseases))

#this is creating the combined data frame
combined_dictionary <- data.frame(
  Other_Columns = c(clean_other_cols, rep("", max_ln_all - length(clean_other_cols))),
  Disease_Causes = c(clean_diseases_cols, rep("", max_ln_all - length(clean_diseases_cols))),
  Non_Disease_Causes = c(clean_non_diseases, rep("", max_ln_all - length(clean_non_diseases)))
)

#this is to create a table to show both diseases and non-diseases but also the other columns
max_ln <- max(length(clean_diseases_cols) , length(clean_non_diseases))

comparison_table <- data.frame(
  Disease_Causes = c(clean_diseases_cols, rep(NA, max_ln - length(clean_diseases_cols))),
  Non_Disease_Causes = c(clean_non_diseases, rep(NA, max_ln - length(clean_non_diseases)))
)

comparison_table[is.na(comparison_table)] <- ""

#this create a better looking Table
combined_dictionary %>%
  kbl(col.names = c("Data Columns", "Diseases Columns", "Non-Disease Columns"),
      caption = "Dataset Column Classification",
      align = "lll") %>%
  kable_styling(bootstrap_options = c("bordered", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 26) %>%
  # Header Styling
  row_spec(0, background = "steelblue", color = "white", bold = TRUE) %>%
  # Column Boxing and Widths
  column_spec(1, border_left = TRUE, border_right = TRUE, width = "15em", bold = TRUE) %>%
  column_spec(2, border_right = TRUE, width = "18em") %>%
  column_spec(3, border_right = TRUE, width = "18em") %>% 
  scroll_box(width = "1400px", height = "550px")
```

------------------------------------------------------------------------

::: {style="float: left; width: 48%; padding-top: 10px;"}
### **The Data Collaboration**

The **"Cause of Deaths"** dataset provides the *counts*, but the **"World Population"** dataset provides the *context*.

By merging them, we move from raw totals to meaningful **rates**. This allows us to answer not just *"how many died?"*, but **"how heavily was the population actually impacted?"**

<br>

### **The Investment Perspective**

**Imagine we are a global healthcare fund.** How do we choose where to invest? 

We don't just look at who has the most deaths; we analyze:
- **The Crude Death Rate:** The true burden on the people.
- **The Trend:** Is the crisis improving or worsening?
:::

::: {style="float: right; width: 45%; background-color: #F0F8FF; padding: 10px; border-radius: 12px; border: 2px solid #4682B4; color: #191970;"}
### **Our Core Questions**

- **Regional Burden:** Which continents face the heaviest impact from biological diseases?


- **Fair Comparison:** How do we level the playing field between small nations and global giants?


- **Trends Over Time:** How has the relationship between population and mortality evolved?


- **Global Reach:** Which specific diseases are the most widespread across international borders?
:::
------------------------------------------------------------------------

```{r echo=FALSE, warning = FALSE}
## Continents with the Most Deaths

dth_pplt_ds %>% 
  select(-c(non_diseases, Total.Death, Crude.Death.Rate)) -> ntrl_deaths

ntrl_deaths %>% 
  mutate(Total.Deaths = rowSums(ntrl_deaths[, 4:22], na.rm = TRUE), 
         Crude.Death.Rate = (Total.Deaths/Population)/100) -> ntrl_deaths

```

---

```{r echo=FALSE}
table_summary <- ntrl_deaths %>% 
  group_by(Continent) %>% 
  summarise(Highest_Death_Count = max(Total.Deaths, na.rm = TRUE)) %>%
  arrange(desc(Highest_Death_Count)) # Sorts highest to lowest

#creates a nice table
table_summary %>%
  mutate(Highest_Death_Count = scales::comma(Highest_Death_Count)) %>%
  kbl(col.names = c("Continent", "Highest Single-Year Total"),
      caption = "Peak Natural Mortality Records by Continent") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = F, 
                position = "left") %>%
  # Header: Navy Blue with White Text
  row_spec(0, background = "midnightblue", color = "white", bold = TRUE) %>%
  # Highlight the Data Column: Light Blue background
  column_spec(2, background = "#F0F7FB", color = "midnightblue", bold = TRUE) %>%
  # Add a border to separate columns
  column_spec(1, border_right = TRUE, bold = TRUE)

#a more better looking bar graph
ntrl_deaths %>% 
  ggplot(aes(x = reorder(Continent, Total.Deaths, FUN = median), 
             y = Total.Deaths, 
             fill = factor(Year))) + # This gives each year a different color
  geom_boxplot(alpha = 0.7, outlier.color = "red") + 
  scale_y_log10(labels = scales::comma) + 
  facet_wrap(~Year) + 
  theme_bw() +
  labs(title = "Natural Deaths by Continent", 
       subtitle = "Color-coded by Year",
       y = "Deaths (Log Scale)", x = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") # Hides the legend since the facet labels already show the year


```


### Continents with the Most Deaths

If we look at the graph, we notice that the continent of Asia has the most death throughout the 4 years.

### Why is this?

Lets look deeper as to why this is.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ntrl_deaths %>% group_by(Continent) %>% 
  summarise(Average.Death.Rate = mean(Crude.Death.Rate))

```

The data here shows that the average death rate in higher in Africa. But how is this when there are morepe total deaths in Asia. To understand this, lets learn about the values

-   what is crude death rate?
-   what does it mean for Asia and their deaths?
-   How can we choose between these 2 continents?

---

## Regional Burden

### Which continents face the heaviest impact from biological diseases?

```{r, echo=FALSE}
dis_pop_ds <- dth_pplt_ds %>%  
  mutate(Total.Disease.Deaths = rowSums(across(all_of(disease_cols)))) %>% 
  select(-c(non_diseases, Code, Rank, CCA3, Capital, Density..per.km.., Growth.Rate, World.Population.Percentage, Area..km..))

dis_pop_ds <- dis_pop_ds %>% 
  relocate(Continent, .after = Country.Territory) %>% 
  relocate(Total.Disease.Deaths, .after = Crude.Death.Rate)

#disease crude death rate
dis_crude_rate <- dis_pop_ds %>%
  group_by(Continent) %>%
  summarize(
    # name it clearly to distinguish it from the overall Crude Death Rate
    Avg_Dis_Death_Rate = mean((Total.Disease.Deaths / Population) * 100000, na.rm = TRUE)
  ) %>%
  arrange(desc(Avg_Dis_Death_Rate))
```

```{r, echo=FALSE}
# --- STEP 1: DATA PROCESSING (SILENT) ---
# We save this to 'dis_pop_ds' so it doesn't print a tibble
dis_pop_ds <- dth_pplt_ds %>%  
  mutate(Total.Disease.Deaths = rowSums(across(all_of(disease_cols)))) %>% 
  select(-c(all_of(non_diseases), Code, Rank, CCA3, Capital, 
            matches("Density|Growth|Percentage|Area")))

# Prepare the plot data object (SILENT)
plot_data_100k <- dis_pop_ds %>%
  group_by(Continent, Year) %>%
  summarize(
    Avg_Dis_Rate = mean((Total.Disease.Deaths / Population) * 100000, na.rm = TRUE),
    .groups = "drop"
  )
```
::: {.columns}

::: {.column width="65%"}
```{r, echo=FALSE}
# --- STEP 2: GRAPH RENDER ---
ggplot(plot_data_100k, aes(x = reorder(Continent, -Avg_Dis_Rate), y = Avg_Dis_Rate)) +
  geom_col(fill = "steelblue", color = "white") +
  geom_text(aes(label = round(Avg_Dis_Rate, 0)), 
            vjust = -0.5, 
            size = 2.8) + 
  facet_wrap(~Year, ncol = 4) + 
  expand_limits(y = max(plot_data_100k$Avg_Dis_Rate) * 1.2) +
  theme_minimal() +
  labs(title = "Average Disease-Specific Death Rate per 100,000 People",
    x = "Continent",
    y = "Deaths per 100k Population") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.background = element_rect(fill = "steelblue"),
    strip.text = element_text(color = "white", face = "bold"),
    panel.spacing = unit(1.2, "lines")
  )
```
:::

::: {.column width="35%"}
<style>
.table {
    margin-top: 30px; /* Moves table down from the header */
    border: 2px solid #4682B4; /* Adds a Steelblue border around the whole table */
}
</style>
```{r, echo=FALSE}
# --- STEP 3: TABLE RENDER ---
dis_pop_ds %>%
  group_by(Continent) %>%
  summarize(
    Avg_Rate = mean((Total.Disease.Deaths / Population) * 100000, na.rm = TRUE)
  ) %>%
  arrange(desc(Avg_Rate)) %>%
  kable(col.names = c("Continent", "Avg. Rate"), 
        digits = 2) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = FALSE,       # This shortens the width
    position = "left",        # Aligns it to the left to leave room for text
    font_size = 18            # Increasing font size makes the table "taller"
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "steelblue") %>%
  row_spec(1, bold = TRUE, background = "#F0F8FF")%>%
  # CUSTOMIZE WIDTH HERE:
  # Adjust 'width' (e.g., "150px" or "3in") to find your "Goldilocks" size
  column_spec(1, width = "180px") %>% 
  column_spec(2, width = "180px") %>% 
  row_spec(0, bold = TRUE, color = "white", background = "steelblue") %>%
  row_spec(1, bold = TRUE, background = "#F0F8FF") %>%
  # Optional: Adds vertical spacing to rows to make it even taller
  row_spec(1:6, extra_css = "padding-top: 10px; padding-bottom: 10px;")
```
<div style="background-color: #ffffff; padding: 15px; border-left: 5px solid #191970;color: #191970; font-size: 0.9em; margin-top: 15px; box-shadow: 2px 2px 5px rgba(0,0,0,0.1);">
We show deaths per 100,000 people so that large countries don't overshadow smaller ones. This reveals when major outbreaks or successful medical breakthroughs shifted the numbers.
</div>
:::

:::
---

## Fair Comparison

### How do we level the playing field between small nations and global giants?

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ntrl_deaths %>% ggplot(aes(x = Population, 
                           y = Total.Deaths, 
                           color = Continent)) + 
  geom_point() + scale_x_log10() + scale_y_log10() + facet_wrap(~Year)
```

### Based on this graph, what conclusion can we make?

The real reason why Asia has the highest total deaths. This graph says "more people equals more deaths". Whats the easiest conclusion for Asia?

Asia has more people overall, so its a given that theyll have more deaths overall.

### Death Rates Over Time Based in Continents

```{r, echo = FALSE, warning = FALSE, message = FALSE}
ntrl_deaths %>% 
  group_by(Continent, Year) %>% 
  summarise(Year, 
            Average.Death.Rate = mean(Crude.Death.Rate)) -> Avg.DthRt.Continent

Avg.DthRt.Continent %>% ggplot(aes(x = Year, 
                                   y = Average.Death.Rate, 
                                   color = Continent)) + 
                                 geom_line() + labs(title = "Death Rates Throughout the Years", x = "Year", y = "Death rates per 1000")
```

We could choose Asia based on their total deaths, but...?

what about Africa?

Which continent should we actually choose?
---

## [Insert title]

```{r}
disease_country_count <- death_ds %>%
  pivot_longer(
    cols = -c(Country.Territory, Code, Year, Total.Death),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  filter(Deaths > 0) %>%
  group_by(Disease) %>%
  summarize(
    Countries.Affected = n_distinct(Country.Territory),
    .groups = "drop"
  ) %>%
  arrange(desc(Countries.Affected))

disease_country_count
```

The above information shows which diseases effected the most countries. Below I found which diease effected the most countries and which effected the least.This could help by putting this into anotther study to find which disease should possoibly be researched more to find a possible cure.

```{r}
most_affected <- disease_country_count %>%
  slice_max(Countries.Affected, n = 1)

least_affected <- disease_country_count %>%
  slice_min(Countries.Affected, n = 1)

most_affected
least_affected
```

```{r}
disease_totals <- death_ds %>%
  pivot_longer(
    cols = -c(Country.Territory, Code, Year, Total.Death),
    names_to = "Disease",
    values_to = "Deaths"
  ) %>%
  group_by(Disease) %>%
  summarize(
    Total.Deaths = sum(Deaths, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(Total.Deaths > 0) %>%
  arrange(desc(Total.Deaths))
```

```{r}
disease_totals %>% arrange(desc(Total.Deaths)) %>% slice(1)
disease_totals %>% arrange(Total.Deaths) %>% slice(1)
```
---

## Trends Over Time

### How has the relationship between population and mortality evolved?

---

## Global Reach {data-background="#F0F8FF"}

### Which specific diseases are the most widespread across international borders?

```{r, echo=FALSE}
# Calculate global mean for each disease
global_reach <- dth_pplt_ds %>%
  # We sum all deaths for a disease globally, then divide by total global population
  summarize(across(all_of(disease_cols), 
                   ~ (sum(. , na.rm = TRUE) / sum(Population, na.rm = TRUE)) * 100000)) %>%
  pivot_longer(cols = everything(), 
               names_to = "Disease", 
               values_to = "Global_Avg_Rate") %>%
  arrange(desc(Global_Avg_Rate)) %>%
  slice_head(n = 6)

# Prepare data for the Dot Plot
top_diseases <- global_reach$Disease

dot_plot_data <- dth_pplt_ds %>%
  select(Continent, all_of(top_diseases), Population) %>%
  pivot_longer(cols = all_of(top_diseases), names_to = "Disease", values_to = "Deaths") %>%
  group_by(Continent, Disease) %>%
  # Weighted calculation: Total deaths in continent / Total population in continent
  summarize(Avg_Rate = (sum(Deaths, na.rm = TRUE) / sum(Population, na.rm = TRUE)) * 100000, 
            .groups = "drop") %>%
  mutate(Disease = gsub("\\.", " ", Disease))
```
::: {.columns}

::: {.column width="40%"}
```{r, echo=FALSE}
# Re-using your global_reach processing from before
global_reach %>%
  mutate(Disease = gsub("\\.", " ", Disease)) %>%
  kable(col.names = c("Disease", "Avg. Rate (100k)"), digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 16) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4682B4")
```
<div style="background-color: #ffffff; padding: 15px; border-left: 5px solid #4682B4; color: #191970; font-size: 0.9em; box-shadow: 2px 2px 5px rgba(0,0,0,0.05);">
<b>The "Intensity Gap":</b> The grey lines represent the disparity between regions. A longer line indicates a disease that is highly sensitive to local infrastructure and <b>resource allocation.</b>
</div>
:::

::: {.column width="60%"}
```{r, echo=FALSE}
# Render Dot Plot with high-distinction colors
ggplot(dot_plot_data, aes(x = Avg_Rate, y = reorder(Disease, Avg_Rate), color = Continent)) +
  geom_line(aes(group = Disease), color = "grey85", linewidth = 1.2) +
  geom_point(size = 5, alpha = 0.9) +
  # Using 'Set2' for much better distinction between 6 continents
  scale_color_brewer(palette = "RdYlBu") +
  theme_minimal() +
  labs(title = "The Global Health Gap",
       subtitle = "Average deaths per 100,000 people",
       x = "Deaths per 100k", y = NULL) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", color = "#191970", size = 18),
    axis.text = element_text(size = 11, color = "#333333"),
    legend.text = element_text(size = 10)
  )
```
:::

:::

---